# R Programming for Bioinformatics

## 2.1 Why R?

R is a powerful language for:

* Statistical analysis
* Data visualization
* Bioinformatics workflows

---

## 2.2 RStudio Environment

This section is just for understanding the installation and purpose:
* Download R from CRAN: https://cran.r-project.org/
* Download RStudio from: https://posit.co/download/rstudio-desktop/

Key panels:

* Console: Main command interface
* Script: For writing and saving code
* Environment: Shows loaded objects
* Plots: For viewing generated visualizations

> **Figure Placeholder**: RStudio interface

---

## 2.3 Basic Syntax

```{r}
GeneExpression <- 5.6
Species <- "Homo sapiens"

x <- c(1,2,3,4,5)
mean(x)
```

---

## 2.4 Biological Example

```{r}
GeneExpressions <- c(7.2, 8.5, 6.3, 9.0, 7.8)
mean(GeneExpressions)
```

---

## 2.5 Data Structures

### Vector

```{r}
v <- c(1,2,3)
```

### Data Frame

```{r}
df <- data.frame(Gene=c("A","B"), Expression=c(2.3,3.5))
```

---

## 2.6 Visualization

```{r}
library(ggplot2)

df <- data.frame(Sample=1:5, Expression=GeneExpressions)
ggplot(df, aes(Sample, Expression)) +
  geom_line() +
  geom_point()
```

---

## 📌 Suggested Figures

* RStudio interface
* Example plot output
