# R Programming for Bioinformatics

## Why R?

First of all, R is an open source and widely used programming language. Specifically, R is a powerful language for:

* Statistical analysis
* Data visualization
* Bioinformatics workflows

---

## RStudio Environment

This section is just for understanding the installation and purpose:
* Download R from CRAN: https://cran.r-project.org/
* Download RStudio from: https://posit.co/download/rstudio-desktop/

Key panels:

* Console: Main command interface
* Script: For writing and saving code
* Environment: Shows loaded objects
* Plots: For viewing generated visualizations

> **Figure Placeholder**: RStudio interface

---


## Basic Syntax

```{r}
# Assigning variables
GeneExpression <- 5.6
Species <- "Homo sapiens"

# Numeric vector
x <- c(1,2,3,4,5)

# Basic operations
mean(x)
sum(x)
length(x)
```


## Working with Biological Data

```{r}
# Gene expression values
GeneExpressions <- c(7.2, 8.5, 6.3, 9.0, 7.8)

# Summary statistics
mean(GeneExpressions)
median(GeneExpressions)
sd(GeneExpressions)
```


## Data Structures in R

### Vector

```{r}
v <- c(1,2,3,4,5)
v
```


### Matrix

```{r}
mat <- matrix(1:9, nrow=3, ncol=3)
mat
```



### List

```{r}
my_list <- list(
  Gene="TP53",
  Expression=7.5,
  Active=TRUE
)
my_list
```



### Data Frame

```{r}
df <- data.frame(
  Gene=c("TP53","BRCA1","EGFR"),
  Expression=c(7.2, 8.5, 6.3)
)
df
```



## Importing Data

```{r eval=FALSE}
# Read CSV file
data <- read.csv("gene_expression.csv")

# View first few rows
head(data)
```


## Data Manipulation

```{r}
# Subsetting
df$Gene
df$Expression

# Filtering
df[df$Expression > 7, ]

# Adding new column
df$HighExpression <- df$Expression > 7
df
```



## Control Structures

### If-Else

```{r}
value <- 8

if(value > 7){
  print("High expression")
} else {
  print("Low expression")
}
```


### Loop

```{r}
for(i in 1:5){
  print(i)
}
```

---

## Apply Functions

```{r}
# Using apply
mat <- matrix(1:9, nrow=3)

apply(mat, 1, sum)  # Row sums
apply(mat, 2, mean) # Column means
```

---

## Basic Statistics

```{r}
# Statistical tests
t.test(GeneExpressions)

# Correlation
x <- c(1,2,3,4,5)
y <- c(2,4,6,8,10)
cor(x,y)
```

---

## Visualization

```{r}
library(ggplot2)

df_plot <- data.frame(
  Sample=1:5,
  Expression=GeneExpressions
)

ggplot(df_plot, aes(x=Sample, y=Expression)) +
  geom_line() +
  geom_point() +
  ggtitle("Gene Expression Trend")
```

---

## Boxplot Visualization

```{r}
ggplot(df, aes(x=Gene, y=Expression)) +
  geom_boxplot() +
  ggtitle("Expression Distribution")
```

---

## Biological Example: Comparing Gene Expression

```{r}
control <- c(5.2, 5.5, 5.8, 6.0, 5.6)
treated <- c(7.1, 7.5, 7.8, 8.0, 7.6)

# Combine into data frame
exp_data <- data.frame(
  Expression=c(control, treated),
  Group=rep(c("Control","Treated"), each=5)
)

# Statistical comparison
t.test(Expression ~ Group, data=exp_data)
```

---

## Visualization of Comparison

```{r}
ggplot(exp_data, aes(x=Group, y=Expression, fill=Group)) +
  geom_boxplot() +
  ggtitle("Control vs Treated Expression")
```

---

## Saving Output

```{r eval=FALSE}
# Save data
write.csv(df, "output.csv", row.names=FALSE)

# Save plot
ggsave("plot.png")
```

---

## Summary

In this tutorial, you learned:

* Basic R syntax
* Data structures (vector, matrix, list, data frame)
* Data manipulation techniques
* Statistical analysis
* Visualization using ggplot2
* Application in biological data analysis

---



