
# From Genes to Networks

Biological systems are interconnected. Instead of studying genes in isolation, network biology allows us to understand how proteins interact and coordinate cellular processes.

In this tutorial, we begin with a small set of biologically important genes and expand them into a large, meaningful protein–protein interaction (PPI) network using STRING and R.

---

## Starting with a Simple Network

Before working with real data, we construct a small example network.

```{r message=FALSE, warning=FALSE}
library(igraph)

edges <- data.frame(
  from=c("TP53","MYC","EGFR","BRCA1"),
  to=c("MYC","EGFR","MTOR","TP53")
)

g_simple <- graph_from_data_frame(edges, directed=FALSE)

plot(g_simple,
     main="A Simple Gene Interaction Network",
     vertex.size=30,
     vertex.label.cex=0.8)
```

This introduces the idea of nodes (genes) and edges (interactions).

---

## Moving to Real Biological Data

We now use the STRING database to construct a biologically meaningful network.

```{r}
library(STRINGdb)

string_db <- STRINGdb$new(
  version="11.5",
  species=9606,
  score_threshold=200
)
```

We begin with a set of important genes:

```{r}
GeneNames <- c("TP53","MDM2","BRCA1","EGFR","MYC",
               "AKT1","PIK3CA","CDK1","CDK2","RB1")

genes_df <- data.frame(Gene=GeneNames)
```

These genes are mapped to STRING identifiers:

```{r}
mapped <- string_db$map(genes_df,"Gene",removeUnmappedRows=TRUE)

mapped
```

---

## Expanding into a Large Network

A small gene list alone is not enough to produce a rich network. Therefore, we expand it by including interacting partners.

```{r}
library(igraph)

hits <- mapped$STRING_id

# First-level neighbors
neighbors_lvl1 <- string_db$get_neighbors(hits)

# Second-level neighbors (more expansion)
neighbors_lvl2 <- string_db$get_neighbors(neighbors_lvl1)

# Combine all nodes
all_nodes <- unique(c(hits, neighbors_lvl1, neighbors_lvl2))

# Sample to keep network manageable but rich
set.seed(123)
all_nodes <- sample(all_nodes, min(180, length(all_nodes)))

# Retrieve interactions
edges <- string_db$get_interactions(all_nodes)

# Relax filtering to ensure dense network
edges <- edges[edges$combined_score >= 250, ]

# Fallback if too sparse
if(nrow(edges) < 300){
  edges <- string_db$get_interactions(all_nodes)
}

# Build graph
g <- graph_from_data_frame(edges[,c("from","to")], directed=FALSE)
g <- simplify(g)

cat("Number of nodes:", vcount(g), "\n")
cat("Number of edges:", ecount(g), "\n")
```

This guarantees a sufficiently large and visually meaningful network.

---

## Visualizing the Network

We highlight the original query genes within the expanded network.

```{r}
V(g)$color <- ifelse(V(g)$name %in% hits, "tomato", "skyblue")

plot(g,
     main="Expanded STRING Network",
     vertex.size=5,
     vertex.label=NA)
```

---

## Understanding Network Structure

We now quantify the importance of each gene using centrality measures.

```{r}
degVals <- degree(g)
betVals <- betweenness(g)
cloVals <- closeness(g)

centrality_df <- data.frame(
  Gene=V(g)$name,
  Degree=degVals,
  Betweenness=betVals,
  Closeness=cloVals
)

head(centrality_df)
```

#### Interpretation

* **Degree** → identifies hub genes (high connectivity)
* **Betweenness** → identifies bottlenecks controlling flow
* **Closeness** → identifies genes with efficient communication

---

## Highlighting Important Genes

We encode importance visually.

```{r}
V(g)$size <- scales::rescale(degVals, to=c(3,12))
V(g)$color <- heat.colors(length(degVals))[rank(degVals)]

plot(g,
     main="Network Highlighting Hub Genes",
     vertex.label=NA)
```

Nodes that are larger and warmer in color are more influential.

---

## Detecting Functional Modules

Biological networks often contain clusters corresponding to functional groups.

```{r}
clusters <- cluster_louvain(g)

V(g)$color <- clusters$membership

plot(g,
     main="Network Communities",
     vertex.size=5,
     vertex.label=NA)
```

---

## Interactive Network Exploration

Finally, we create an interactive network for deeper exploration.

```{r}
library(visNetwork)

nodes <- data.frame(
  id = V(g)$name,
  label = V(g)$name,
  group = clusters$membership,
  value = degVals
)

edges_df <- as.data.frame(get.edgelist(g))
colnames(edges_df) <- c("from","to")

visNetwork(nodes, edges_df) %>%
  visNodes(scaling=list(min=5, max=30)) %>%
  visEdges(smooth=FALSE) %>%
  visOptions(highlightNearest=TRUE,
             nodesIdSelection=TRUE) %>%
  visLayout(randomSeed=42)
```

---

## Identifying Hub Genes

We identify the most connected genes in the network.

```{r}
hubGenes <- V(g)$name[degVals > quantile(degVals, 0.75)]

cat("Hub Genes Identified:\n")
print(hubGenes)
```

---

